---
title: "Housing prices prediction"
output: html_document
date: "2026-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




## Housing Data set

```{r message=FALSE, warning=FALSE}
# Load required liraries
library(tidyverse)
library(caret)
library(randomForest)
library(MASS)
library(fmsb)

# Load cleaned Housing Data set
housing_df <- read.csv("Housingdata_cleaned.csv")
```


### METHOD 1: Recursive Feature Elimination (RFE)

```{r}
#Control
control <- rfeControl(functions = rfFuncs,
                      method = "cv",
                      number = 5)

# set random seed
set.seed(123)

rfe_model <- rfe(
  x = housing_df[, names(housing_df) != "SalePrice"],
  y = housing_df$SalePrice,
  sizes = c(5,10,15,20),
  rfeControl = control
)
```

```{r}
#Selected Features
selected_features_rfe <- predictors(rfe_model)

print("Selected Features using RFE:")
print(selected_features_rfe)

```

```{r}
# Feature Importance Plot
plot(rfe_model, type = c("g", "o"))

```

### METHOD 2: Stepwise Feature Selection (Forward + Backward)

embedded method using AIC to select the best model.

```{r}
# Full Model
full_model <- lm(SalePrice ~ ., data = housing_df)
# Null Model
null_model <- lm(SalePrice ~ 1, data = housing_df)

# Stepwise Selection
step_model <- stepAIC(
  null_model,
  scope = list(lower = null_model, upper = full_model),
  direction = "both",trace = FALSE)
```


```{r}
# Selected Features
selected_features_step <- names(coef(step_model))[-1]

print("Selected Features using Stepwise Selection:")
print(selected_features_step)
```
# Part 2: Model Training and Comparison 

## 1. Train Two Models 

```{r}
# Feature selected by both feature selection methodology

# Recursive Feature Elimination (RFE)
features_rfe <- c(
"GrLivArea","OverallQual","TotalBsmtSF","X1stFlrSF","GarageCars",
"X2ndFlrSF","BsmtFinSF1","GarageArea","YearBuilt","MSZoning",
"LotArea","ExterQual","Fireplaces","YearRemodAdd","GarageFinish",
"OverallCond","CentralAir","MSSubClass","BsmtFinType1","Neighborhood"
)

# Stepwise Feature Selection
features_stepwise <- c(
"OverallQual","GrLivArea","Neighborhood","BsmtFinSF1","MSSubClass",
"OverallCond","YearBuilt","LotArea","TotalBsmtSF","Fireplaces",
"MasVnrArea","GarageCars","YearRemodAdd","TotRmsAbvGrd"
)

```


```{r}
# Train/Test Split - 80% train data and 20% test data

#set random seed
set.seed(123)
trainIndex <- createDataPartition(housing_df$SalePrice, p=0.8, list=FALSE)
train_data <- housing_df[trainIndex,]
test_data <- housing_df[-trainIndex,]

```


### Linear Regression Models


```{r}
# Model using RFE features
formula_rfe <- as.formula(
paste("SalePrice ~", paste(features_rfe, collapse="+"))
)
# Train Linear regression model
lm_rfe <- lm(formula_rfe, data=train_data)
# Prediction on test data
pred_lm_rfe <- predict(lm_rfe, test_data)

```


```{r}
# Model using Stepwise features
formula_step <- as.formula(
paste("SalePrice ~", paste(features_stepwise, collapse="+"))
)

# Train Linear regression model
lm_step <- lm(formula_step, data=train_data)
# Prediction on test data
pred_lm_step <- predict(lm_step, test_data)
```


### Random Forest Models

```{r}
# Train Linear regression model - RFE features
rf_rfe <- randomForest(
formula_rfe,
data=train_data,
ntree=200
)
# Prediction on test data
pred_rf_rfe <- predict(rf_rfe, test_data)

```


```{r}
# Train Linear regression model - Stepwise features
rf_step <- randomForest(
formula_step,
data=train_data,
ntree=200
)
# Prediction on test data
pred_rf_step <- predict(rf_step, test_data)
```


## 2. Evaluation and Compare

```{r}

# Metrics for evaluation
metrics <- function(actual,pred){

MAE <- mean(abs(actual-pred))
MSE <- mean((actual-pred)^2)
RMSE <- sqrt(MSE)
R2 <- cor(actual,pred)^2

return(c(MAE,MSE,RMSE,R2))

}

```

```{r}
# Evaluation of each model
m1 <- metrics(test_data$SalePrice,pred_lm_rfe)
m2 <- metrics(test_data$SalePrice,pred_lm_step)
m3 <- metrics(test_data$SalePrice,pred_rf_rfe)
m4 <- metrics(test_data$SalePrice,pred_rf_step)

results <- data.frame(
Model=c("LM-RFE","LM-STEP","RF-RFE","RF-STEP"),
MAE=c(m1[1],m2[1],m3[1],m4[1]),
MSE=c(m1[2],m2[2],m3[2],m4[2]),
RMSE=c(m1[3],m2[3],m3[3],m4[3]),
R2=c(m1[4],m2[4],m3[4],m4[4])
)

print(results)
```


## 3. Visualizations 

### Spider Chart

```{r}
# 
radar_data <- results[,2:5]
radar_data <- rbind(apply(radar_data,2,max),
                    apply(radar_data,2,min),radar_data)

rownames(radar_data) <- c("max","min",results$Model)

radarchart(radar_data,axistype=1,title="Model Comparison Metrics")

```

### Residual Plots

```{r}
# Linear Regression
res_lm_rfe  <- test_data$SalePrice - pred_lm_rfe
res_lm_step <- test_data$SalePrice - pred_lm_step

df_lm_res <- data.frame(
Predicted = c(pred_lm_rfe, pred_lm_step),
Residuals = c(res_lm_rfe, res_lm_step),
Method = rep(c("RFE Features","Stepwise Features"), each = length(test_data$SalePrice)))

ggplot(df_lm_res, aes(x = Predicted, y = Residuals, color = Method)) + 
  geom_point(alpha = 0.6) + theme_bw() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot - Linear Regression",
       x = "Predicted SalePrice", y = "Residuals")
```

```{r}
# Random Forest
res_rf_rfe  <- test_data$SalePrice - pred_rf_rfe
res_rf_step <- test_data$SalePrice - pred_rf_step

df_rf_res <- data.frame(
Predicted = c(pred_rf_rfe, pred_rf_step),
Residuals = c(res_rf_rfe, res_rf_step),
Method = rep(c("RFE Features","Stepwise Features"), each = length(test_data$SalePrice))
)

ggplot(df_rf_res, aes(x = Predicted, y = Residuals, color = Method)) +
  geom_point(alpha = 0.6) + theme_bw() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residual Plot - Random Forest",
       x = "Predicted SalePrice", y = "Residuals")
```

### Predicted vs Actual Scatter Plot

```{r}
# Linear Regression
df_lm <- data.frame(
Actual = test_data$SalePrice,
Predicted = c(pred_lm_rfe, pred_lm_step),
Method = rep(c("RFE Features","Stepwise Features"), each=length(test_data$SalePrice)))

ggplot(df_lm, aes(x=Actual, y=Predicted, color=Method)) +
  geom_point(alpha=0.6) + theme_bw() +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  labs( title="Predicted vs Actual SalePrice (Linear Regression)",
        x="Actual SalePrice",y="Predicted SalePrice")

```
```{r}
# Random Forest
df_rf <- data.frame(
Actual = test_data$SalePrice,
Predicted = c(pred_rf_rfe, pred_rf_step),
Method = rep(c("RFE Features","Stepwise Features"), each=length(test_data$SalePrice))
)

ggplot(df_rf, aes(x=Actual, y=Predicted, color=Method)) +
  geom_point(alpha=0.6) + theme_bw() +
  geom_abline(slope=1, intercept=0, linetype="dashed") +
  labs( title="Predicted vs Actual SalePrice (Random Forest)",
        x="Actual SalePrice",y="Predicted SalePrice")
```

