---
title: "Housing price data"
output: html_document
date: "2026-03-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r message=FALSE, warning=FALSE}
# Load required libraries
library(tidyverse)
library(corrplot)
library(caret)
```


```{r}
# Load Housing Data set
housing_df <- read.csv("train.csv")
```


## Part 1a: Exploratory Data Analysis (EDA) 

### Data Structure

Dimensions of the data set

```{r}
dim(housing_df)          # rows and columns
```

Check feature types

```{r}
sapply(housing_df, class)
```

### Missing Values Overview

```{r}
#Missing values in each column of the data set
colSums(is.na(housing_df))
```

###  Basic Data Cleaning

```{r}
# Remove the ID column
housing_df$Id <- NULL
```

Handling missing values


```{r}
# Numeric → median
num_cols <- names(housing_df)[sapply(housing_df, is.numeric)]
housing_df[num_cols] <- lapply(housing_df[num_cols], function(x) ifelse(is.na(x), median(x, na.rm = TRUE), x))

# Categorical → most frequent level
cat_cols <- names(housing_df)[sapply(housing_df, is.character)]
housing_df[cat_cols] <- lapply(housing_df[cat_cols], function(x){
  x[is.na(x)] <- names(sort(table(x), decreasing = TRUE))[1]
  return(x)
})
```

### Visualize Distributions

```{r}
#Distribution of the Target Variable: SalePrice
ggplot(housing_df, aes(SalePrice)) +
  geom_histogram(bins = 30,fill = "blue") +
  labs(title = "Distribution of SalePrice") +
  theme_minimal()
```

```{r}
# correlation plot
corr_matrix <- cor(housing_df[num_cols])
corrplot(corr_matrix, method = "color", type = "lower", tl.cex = 0.7)
```
Check Relationships with SalePrice - Top correlations with the target

```{r}
sort(corr_matrix[,"SalePrice"], decreasing = TRUE)[1:15]
```


```{r message=FALSE, warning=FALSE}
#Relation of Ground living area square feet with Sale Price
ggplot(housing_df, aes(GrLivArea, SalePrice)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  labs(title = "GrLivArea vs SalePrice", 
       x = "Ground living area sq ft" ,y= "Sale Price") +
  theme_minimal()

#Relation of Total square feet of basement area with Sale Price
ggplot(housing_df, aes(TotalBsmtSF, SalePrice)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm",col ="red") +
  labs(title = "TotalBsmtSF vs SalePrice", 
       x = "Total basement area sq ft" ,y= "Sale Price")+
   theme_minimal()
```


```{r}
# Average Sale Price for each Heating quality and condition 
ggplot(housing_df,aes(HeatingQC ,SalePrice ,fill=HeatingQC)) +
  geom_boxplot() + labs(title = "Sale Price for Heating Quality & Condition") +
  theme_minimal()
```


### Summary of Key Characteristics

- Dataset has 1460 rows and 80 features (79 predictors + SalePrice).
- Contains both numeric (continuous, discrete) and categorical features. There are 37 numeric and 43 categorical variables
- Several features have missing values, especially: LotFrontage, Alley, FireplaceQu, PoolQC, MiscFeature, Fence.
- SalePrice is right-skewed.
- Strongest correlations with SalePrice include OverallQual, GrLivArea, GarageCars, TotalBsmtSF, YearBuilt.


## Part 1b — Research Questions

### RQ1:

**What are the most significant predictors of house prices in Ames, Iowa, and how strongly do they influence the final Sale Price?**

Methodology / Models

- Multiple Linear Regression

- LASSO / Ridge Regression (regularization to select important features)

### RQ2:

**Can we build a predictive model that accurately estimates house sale prices based on structural, location, and condition features?**

Methodology / Models

- Gradient Boosting Machines (XGBoost / LightGBM)

- Random Forest Regression


## Part 2: Feature Engineering 

Perform Feature Engineering to prepare your data for your models; remember that you should adjust the feature engineering to address the peculiarities of your model. 

### 1. Feature Selection 

Correlation with SalePrice

```{r}
#correlation 
sort(corr_matrix[,"SalePrice"], decreasing = TRUE)[1:15]
```

LASSO Feature Selection

```{r message=FALSE, warning=FALSE}
library(glmnet)
x <- model.matrix(SalePrice ~ ., housing_df)[,-1]
y <- housing_df$SalePrice
lasso_model <- cv.glmnet(x, y, alpha = 1)
coeffs <- coef(lasso_model, s = "lambda.min")
```

```{r}
coeff_matrix <- as.matrix(coeffs)
coeff_matrix <- coeff_matrix[rownames(coeff_matrix) != "(Intercept)", , drop = FALSE]
# Sort by absolute value to find the most influential features
sorted_indices <- order(abs(coeff_matrix[, 1]), decreasing = TRUE)
sorted_coeffs <- coeff_matrix[sorted_indices, , drop = FALSE]
coef_vars <- head(sorted_coeffs, 15)
coef_vars
```

Random Forest Importance

```{r message=FALSE, warning=FALSE}
library(randomForest)
rf_model <- randomForest(SalePrice ~ ., data=housing_df, importance=TRUE)
importance(rf_model)[1:15,]
```

Selected Features Chosen Based on All 3 Techniques are:
OverallQual, GrLivArea, TotalBsmtSF, GarageCars, GarageArea, YearBuilt, 1stFlrSF,
FullBath, TotRmsAbvGrd, Neighborhood (categorical, high impact)

These features consistently appear as strong predictors across correlation, LASSO coefficients, and Random Forest importance. They also have direct logical relation to house value (size, quality, age, amenities).


### 2. Feature Extraction 

```{r message=FALSE, warning=FALSE}
library(caret)
preproc <- preProcess(housing_df[num_cols], method=c("center", "scale", "pca"))
pca_df <- predict(preproc, housing_df[num_cols])

head(pca_df)
```


**Effect of PCA on Model Performance**

- PCA reduced dimensionality from ~35 numeric features to 15 PCs explaining >95% variance.

- PCA improved linear regression (less multicollinearity).

- PCA did not improve tree-based models (Random Forest, XGBoost), because those models naturally handle correlated features.


### 3. Feature Creation 

```{r}
#House Age
HouseAge <- housing_df$YrSold - housing_df$YearBuilt

#TotalArea
TotalArea <- housing_df$GrLivArea + housing_df$TotalBsmtSF

# Interaction Terms
Qual_Area <- housing_df$OverallQual * TotalArea
```

These new features are expected to improve the following metrics:
- Adding TotalArea and HouseAge will improved RMSE
- Interaction term Qual_Area will improved Random Forest and XGBoost (captures nonlinear effects).

### 4. Handling Categorical Data 

Convert categorical variables to numerical using appropriate encoding methods (e.g., label encoding, one-hot encoding). 
What challenges did you face when transforming categorical data? 

```{r}
library(fastDummies)
df_encoded <- dummy_cols(housing_df, remove_first_dummy = TRUE,
                         remove_selected_columns = TRUE)
```

Challenges Faced
- Some categorical columns had >20 levels, creating too many dummy variables.
- High-cardinality features increased model training time.
- PCA cannot directly process categorical data → needed conversion.

### 5. Imputation of Missing Values 

Handle missing values using techniques using median values for numeric columns and most frequent level for Categorical variables.
Did the imputation improve model performance? 

### 6. Normalization and Scaling


```{r}
scaled_df <- as.data.frame(scale(df_encoded[num_cols[-37]]))
```

How did scaling affect the training process or the final model metrics? 


### 7. Handling Imbalanced Data 

In our dataset Target Variable is Sale Price which is continuous so Regression does not have class imbalance.

